---
output: 
  html_document: 
    keep_md: yes
---
Assignment: Course Project 1
============================

This is R Markdown documentation for course Assignment.

## Loading and preprocessing the data

Here we are loading the Activity data-set in R. We then process the data to create a new data-frame after removing the null "step" records;also we transform the "date" field to R Date data-type.

```{r loading data}
act <- read.csv("/Users/Sandip/Documents/Data/Knowledge\ Base/Big\ Data/R\ Programming/Assignment_Data/activity.csv", header = TRUE, na.strings = "NA")
act_s <- act[!is.na(act$steps),]
act_s$date = as.Date(act_s$date)
summary(act_s)
```

## What is mean total number of steps taken per day?


We first Calculate the total number of steps taken per day and create histogram and bar-plot for the total number of steps taken each day

```{r Steps Count}
# Calculate the count of steps per day
aggr <- aggregate(steps~date,act_s,sum)
par(mfrow = c(1,2), mar = c(6,4,2,2) )
hist(aggr$steps,xlab = "Steps", main = "Frequency of Steps per Day", col = "red",breaks = 53)
barplot(aggr$steps,names.arg = aggr$date, main = "Step Count per Day", xlab = "Date", ylab = "Steps Count")
```


We calculate the Mean and Median of the total number of steps taken for each day and then plot the same using base plot.

```{r, Steps Count Aggregate}
aggr_mean <- aggregate(steps~date,act_s,mean)
aggr_median <- aggregate(steps~date,act_s,median)
par(mfrow = c(1,2), mar = c(6,4,2,2) )
with(aggr_mean, plot (date,steps,type = "l",main = "Per Day Mean for Steps", xlab = "Date", ylab = "Mean"))
with(aggr_mean, points (date,steps,col = "blue"))
with(aggr_median, plot (date,steps,type = "l",main = "Per Day Median for Steps", xlab = "Date", ylab = "Median"))
with(aggr_median, points (date,steps,col = "Blue"))
```

## What is the average daily activity pattern?

Here we create a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r Steps Mean and Median}
aggr_int_mean <- aggregate(steps~interval,act_s,mean)
with(aggr_int_mean, plot (interval,steps,type = "l",axes = FALSE))
with(aggr_int_mean, points (interval,steps))
axis(side = 1, at = seq(0,2500,50))
axis(side = 2, at = seq(0,210,20))
```


Following 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps:

```{r aggregate}
aggr_int_mean[aggr_int_mean$steps == max(aggr_int_mean$steps),1]
```

## Imputing missing values

We now calculate the total no. of missing values ("NA") in the data-set -
```{r Missing Count}
sapply(act, function(x) sum(is.na(x)))
```

Here we replace "NA"s will mean value for every 5 min interval; also we convert the date field to R Date data-type:

```{r replace missing}
x = aggregate(steps~interval,act,mean,na.rm = TRUE)

for (i in x$interval){
        act[(is.na(act$steps) & (act$interval == x$interval)),1] = x$steps
}
act$date = as.Date(act$date)
```

Inspecting the data-frame again to find if all the "NA"s are removed -

```{r validate}
sapply(act, function(x) sum(is.na(x)))
```

Here we plot the histogram for total number of steps per day
```{r plot hist}
aggr <- aggregate(steps~date,act,sum)
hist(aggr$steps,xlab = "Steps", main = "Frequency of Steps per Day", col = "red",breaks = 60)
```

We calculate the Mean and Median of the total number of steps taken for each day and then plot the same using base plot.

```{r Plot Mean and Median}
aggr_mean <- aggregate(steps~date,act,mean)
aggr_median <- aggregate(steps~date,act,median)
par(mfrow = c(1,2), mar = c(6,4,2,2) )
with(aggr_mean, plot (date,steps,type = "l",main = "Per Day Mean for Steps", xlab = "Date", ylab = "Mean"))
with(aggr_mean, points (date,steps,col = "blue"))
with(aggr_median, plot (date,steps,type = "l",main = "Per Day Median for Steps", xlab = "Date", ylab = "Median"))
with(aggr_median, points (date,steps,col = "Blue"))
```
Here we find that, for the new data set the value of Median is greater than "0" for few days compared to all "0"s in the previous data-set.

## Are there differences in activity patterns between weekdays and weekends?

We now add a new variable to the data-frame to indicate whether the date is a Weekday or Weekend
```{r Creating Factor}
act$Weekday = "Weekday"
act[weekdays(act$date) %in% c("Saturday", "Sunday"),4] = "Weekend"
aggr <- aggregate(steps~interval+Weekday,act,mean)
aggr$Weekday = as.factor(aggr$Weekday)
```

Here we plot Average Number of Steps per 5 min interval for all the days (grouped by Weekday indicator). We use Lattice plotting system here.

```{r Lattice Plot}
library(lattice)
xyplot((aggr$steps)~aggr$interval|aggr$Weekday, type = "l", xlab = "5 Min Interval", ylab = "Average Steps", layout = c(1,2))
```

